---
title: "hw05-gapminder"
output:
  github_document
---

# Factor and figure management

## Initialize the data

* Load the required libraries:

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(broom))
```

* We'll use *forcats* to help re-order factors (package located inside the tidyverse) and *plotly* to enhance the plot visualization. *Broom* is handy for providing statistical analyses.
* We will check the structure of the data first to *sanity check* that we are working with factors:

```{r}
gapminder %>% str()
```

* We've confirmed that *country* is a factor with 142 levels and *continent* is a factor with 5 levels. We'll proceed with manipulating the data.

## Part 1 - Factor Management
*Characterize data before and after re-levelling*

Drop Oceania. Filter the Gapminder data to remove observations associated with the continent of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors.

Reorder the levels of country or continent. Use the forcats package to change the order of the factor levels, based on a principled summary of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the median.

### Method
* Drop Oceania by filtering the data to remove observations
* Remove unused factor levels
* Provide concrete information on the data before/after the manipulation (e.g. effect on number of rows/levels of affected factors)
* Re-order the levels of continent: use forcats to change the order of the factor levels based on a principled summary of one of the quantitative variables (consider something other than median)

### Code
First, we will examine a summary of the initial data:

```{r}
gapminder %>% 
  summary() %>% #Evaluate the factors and observations per factor
  kable() #Enhance the table output
```

Oceania has 24 observations in the original data set. Next we will drop the observations related to Oceania from the data set:

```{r}
gapminder_dropOc <- gapminder %>%
  filter(continent != "Oceania") 

gapminder_dropOc %>% 
  summary() %>%  
  kable() 

gapminder_dropOc %>% 
  str()
```

Oceania now has 0 observations, however it is still present in the data frame as a factor. Next, we will drop Oceania as an unused factor using the *droplevels()* function from the *forcats* package:

```{r}
gapminder_dropOc <- gapminder_dropOc %>%
  droplevels()

gapminder_dropOc %>% 
  summary() %>% 
  kable()       #highlight cell in table if possible

gapminder_dropOc %>% 
  str()
```

We observe that Oceania has now disappeared from the continent list. Further, we see that the factor continent now has only 4 levels. The original data set had 1704 observations of 6 variables (i.e. 1704 rows and 6 columns), while the manipulated data set has 1680 observations of 6 variables. Therefore we observe a concrete reduction on the data. When manipulating a data set through filters, it is advantageous to create a new variable for the manipulated data in order to refer to it in the future and maintain the integrity of the original data set.

* Next we will create a principled summary of the data based on the quantitative variable *gdpPercap*:
```{r}
gdpPercap_ <- gapminder_dropOc %>% 
  select(continent, year, gdpPercap) %>% #Reduce the size of the data set
  group_by(continent) %>% 
  kable()
  #spread(key = "continent", value = "gdpPercap")


```


Reorder the levels of country or continent. Use the forcats package to change the order of the factor levels, based on a principled summary of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the median.

Be sure to also characterize the (derived) data before and after your factor re-leveling:

Explore the effects of arrange(). Does merely arranging the data have any effect on, say, a figure?
Explore the effects of reordering a factor and factor reordering coupled with arrange(). Especially, what effect does this have on a figure?
These explorations should involve the data, the factor levels, and some figures.

```{r}
ggplot(gap_america) +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas",
         x = "Country", y = "Life Expectancy")

gap_america %>%
  mutate(country = fct_reorder(country, pop)) %>% #Reorder factor
  ggplot() +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas (ordered by population)",
         x = "Country", y = "Life Expectancy")

  droplevels() %>%
  group_by(country) %>%
  mutate(mean_pop = mean(pop)) %>%
  arrange(mean_pop) #Arrange by mean poplutation some all country measurements stay

kable(head(gap_america_arr, 15))
```

## Part 2 - File I/O
## Part 3 - Visualization Design
## Part 4 - Writing figures to file
## Bonus - Re-evaluate a factor
