---
title: "hw05-gapminder"
output:
  github_document
---

# Factor and figure management

## Initialize the data

* Load the required libraries:

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(scales))
```

* We'll use *forcats* to help re-order factors (package located inside the tidyverse) and *plotly* to enhance the plot visualization. *Broom* is handy for providing statistical analyses.
* We will check the structure of the data first to *sanity check* that we are working with factors:

```{r}
gapminder %>% str()
```

* We've confirmed that *country* is a factor with 142 levels and *continent* is a factor with 5 levels. We'll proceed with manipulating the data.

## Part 1 - Factor Management
*Characterize data before and after re-levelling*

Drop Oceania. Filter the Gapminder data to remove observations associated with the continent of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors.

Reorder the levels of country or continent. Use the forcats package to change the order of the factor levels, based on a principled summary of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the median.

### Method
* Drop Oceania by filtering the data to remove observations
* Remove unused factor levels
* Provide concrete information on the data before/after the manipulation (e.g. effect on number of rows/levels of affected factors)
* Re-order the levels of continent

### Code
First, we will examine a summary of the initial data:

```{r}
gapminder %>% 
  summary() %>% #Evaluate the factors and observations per factor
  kable() #Enhance the table output
```

Oceania has 24 observations in the original data set. Next we will drop the observations related to Oceania from the data set:

```{r}
gapminder_dropOc <- gapminder %>%
  filter(continent != "Oceania") 

gapminder_dropOc %>% 
  summary() %>%  
  kable() 

gapminder_dropOc %>% 
  str()
```

Oceania now has 0 observations, however it is still present in the data frame as a factor. Next, we will drop Oceania as an unused factor using the *droplevels()* function from the *forcats* package:

```{r}
gapminder_dropOc <- gapminder_dropOc %>%
  droplevels()

gapminder_dropOc %>% 
  summary() %>% 
  kable()     

gapminder_dropOc %>% 
  str()
```

We observe that Oceania has now disappeared from the continent list. Further, we see that the factor continent now has only 4 levels. The original data set had 1704 observations of 6 variables (i.e. 1704 rows and 6 columns), while the manipulated data set has 1680 observations of 6 variables. Therefore we observe a concrete reduction on the data. When manipulating a data set through filters, it is advantageous to create a new variable for the manipulated data in order to refer to it in the future and maintain the integrity of the original data set.

* Next we will create a principled summary of the data based on the quantitative variable *gdpPercap*. We want to evaluate the rank of countries in Europe based on gdpPercap. We will start by plotting the data as is:

```{r}
Europe_gdp <- gapminder %>% 
  select(continent, country, gdpPercap,year) %>%  #Reduce the size of the data set for faster processing
  filter(continent == "Europe") 

Europe_gdp %>% 
  str()

Europe_gdp %>% 
  head() %>% 
  kable()

Europe_gdp %>% 
  ggplot(aes(country, gdpPercap)) + 
  geom_violin() +
  labs(title = "GDP per capita - Europe",
    x = "Country", y = "GDP per capita") +
  scale_y_log10(labels=dollar_format()) +
  #scale_x_log10(labels=comma_format()) + #Apply the scale package to suit the scale to the values
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10))  #Rotate x labels

  
```

Plotting the data directly doesn't provide any insight into a correlation between these two variables. We will next *arrange* by gdpPerCap to see whether this has an effect on the table and plot:

```{r}
Europe_gdp_arr <- Europe_gdp %>% 
  arrange(gdpPercap)

Europe_gdp_arr %>% 
  str()

Europe_gdp_arr %>% 
  head() %>% 
  kable()

Europe_gdp_arr %>% 
  ggplot(aes(country, gdpPercap)) + 
  geom_violin() +
  labs(title = "GDP per capita - Europe",
    x = "Country", y = "GDP per capita") +
  scale_y_log10(labels=dollar_format()) +
  #scale_x_log10(labels=comma_format()) + #Apply the scale package to suit the scale to the values
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10))  #Rotate x labels
```

We observe the *arrange* did not affect the structure or the plot, however the table output is clearly different. Next we will evaluate the effect of using the *forcats* package to re-order the data:

```{r}
Europe_gdp %>% 
  ggplot(aes(fct_reorder(country, gdpPercap), gdpPercap)) + 
  geom_violin() +
  labs(title = "GDP per capita - Europe",
    x = "Country", y = "GDP per capita") +
  scale_y_log10(labels=dollar_format()) +
  #scale_x_log10(labels=comma_format()) + #Apply the scale package to suit the scale to the values
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10))  #Rotate x labels

```

We observe that *fct_reorder* does affect the plot, as it now shows countries in order of ascending median GDP per capita. Finally, we will examine the effect of combining *arrange* and *fct_reorder*:

```{r}

Europe_gdp_arr %>% 
  ggplot(aes(fct_reorder(country, gdpPercap), gdpPercap)) + 
  geom_violin() +
  labs(title = "GDP per capita - Europe",
    x = "Country", y = "GDP per capita") +
  scale_y_log10(labels=dollar_format()) +
  #scale_x_log10(labels=comma_format()) + #Apply the scale package to suit the scale to the values
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10))  #Rotate x labels

```

Thus, coupling *arrange* and *fct_reorder* allows us to manipulate the structure, table and plot, so it is more comprehensive to use both commands if all three types of output are desired.

## Part 2 - File I/O

Experiment with one or more of write_csv()/read_csv() (and/or TSV friends), saveRDS()/readRDS(), dput()/dget(). 
Create something new, probably by filtering or grouped-summarization of Singer or Gapminder. 
I highly recommend you fiddle with the factor levels, i.e. make them non-alphabetical (see previous section). 
Explore whether this survives the round trip of writing to file then reading back in.

### Method
* Create a new data set
* Arrange the data set
* Write to file
* Read from file

### Code
``` {r}
gapminder_lifeExp <- gapminder %>% 
  select(lifeExp, continent, year) %>% 
  group_by(continent, year) %>% 
  summarise(mean_gdp = mean(lifeExp))

gapminder_lifeExp %>% 
  head() %>% 
  kable()

gapminder_lifeExp_arr <- gapminder_lifeExp %>% 
  arrange(mean_gdp)

gapminder_lifeExp_arr %>% 
  head() %>% 
  kable()
```

We observe the difference between the tables after the *arrange* function has been applied. Now we will test if the arrangment is preserved after writing and reading to a file:

```{r}
write_csv(gapminder_lifeExp_arr, "gapminder_lifeExp_arr.csv")

read_csv("gapminder_lifeExp_arr.csv") %>% 
  head() %>% 
  kable()
```

The new arrangement of the data was preserved, therefore we learned that *arrange* can be used to manipulate tables that are written externally to files, in addition to tables internally within the R environment, as we learned in the previous section.

## Part 3 - Visualization Design

Remake at least one figure or create a new one, in light of something you learned in the recent class meetings about visualization design and color. Maybe juxtapose your first attempt and what you obtained after some time spent working on it. Reflect on the differences. If using Gapminder, you can use the country or continent color scheme that ships with Gapminder. Consult the dimensions listed in All the Graph Things.

Then, make a new graph by converting this visual (or another, if you’d like) to a plotly graph. What are some things that plotly makes possible, that are not possible with a regular ggplot2 graph?

```{r}


```

## Part 4 - Writing figures to file

Use ggsave() to explicitly save a plot to file. Then use dfdf to load and embed it in your report. You can play around with various options, such as:

Arguments of ggsave(), such as width, height, resolution or text scaling.
Various graphics devices, e.g. a vector vs. raster format.
Explicit provision of the plot object p via ggsave(..., plot = p). Show a situation in which this actually matters.
